A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-19 podc_paper/algorithm.tex
working on algorithm

We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each node of the tree stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the array in $P$'s leaf.
Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
block and attempts to append this block to $v$'s array using a CAS instruction.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process.
We define a total order on all operations that have been propagated to the root, which is used to
linearize the operations.
In this section, we assume for simplicity that each node has an infinite blocks array.
We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.

@@

>>>>

We begin with an overview of the algorithm.
We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each node of the tree stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the blocks array in $P$'s leaf.
Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
block and attempts to append this block to $v$'s array using a CAS instruction.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process.
We define a total order on all operations that have been propagated to the root, which is used to
linearize the operations.
In this section, we assume for simplicity that each node has an infinite blocks array.
We describe in Section \ref{reducing} how to replace the infinite array by a representation that uses bounded space.

@@

An operation can only terminate after it a block containing it has been appended to the root's blocks array.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Our linearization orders operations according to the block they belong to in the root's blocks array.
Operations that appear in the same block are necessarily concurrent, so they can be ordered arbitrarily.
We next describe how to order them so that computing results of dequeue operations is efficient.

Each block $B$ in a node $v$ represents several consecutive blocks in each of $v$'s children.
The blocks in the children are called the \emph{direct subblocks} of $B$.
A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
or a subblock of a direct subblock of $B$.
Recall that blocks in the leaves contain a single operation.
A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.
We order the operations within a block as follows.

### 2022-12-19 podc_paper/related.tex
working on algorithm

\paragraph{Array-Based Queues}
Arrays can be used to implement queues with bounded capacity, with indices
for the front and back elements updated using CAS instructions 
\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also suffer from the CAS retry problem, with $\Omega(p)$ steps per operation.

@@

\paragraph{Universal Constructions}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
the universal construction of Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(P)$.
\here{Double check this.}

@@

\paragraph{Fetch\&Increment Objects}
Progress has been made on obtaining sublinear-time implementations of other objects 
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increment operations
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.

>>>>

\paragraph{Array-Based Queues}
Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
%Similar to the front and back pointers of list-based queues, indices
Dequeues and enqueues update
indexes of the front and back elements using CAS instructions.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also use $\Omega(p)$ steps per operation due to the CAS retry problem.

@@

\paragraph{Universal Constructions}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
the universal construction of Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(p)$.
\here{Double check this.}

@@

\paragraph{Fetch\&Increment Objects}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.
